---
title: Lab 7 Proposals
authors:
  - name: Miranda Chin
    affiliation: CSU
    roles: writing
    corresponding: true
  - name: Kayla Sutton
    affiliation: CSU
    roles: writing
    corresponding: false
bibliography: references.bib
csl: apa-5th-edition.csl
editor: 
  markdown: 
    wrap: sentence
---

## Ozone pollution associated with Fort Collins policies

*Abstract:*

Air quality is a factor that affects many different facets of life; not only is it an indicator of environmental health and stability, but also affects human health and well being [@fiore2003variability].
Increased rates of ozone has been a growing issue in many different places around the world due to rapidly increasing rates of population, especially in urban areas, as well as behavior such as vehicle use .
Ozone has been proven to be connected to health risks such as asthma and respiratory diseases, putting vulnerable populations at risk as it becomes a more prevalent issue [@kinney2000respiratory]. In this project, we aim to determine if Fort Collins' policies regarding ozone control have been successful in addressing these issues, especially when comparing other counties to Larimer. Our methods are composed of a time series analysis to track ozone over the period of 2003 before policies were passed to 2021 afterwards, as well as a Difference-in-Differences model that will compare Larimer's air quailty index (AQI) to that of Boulder county. We found that while Larimer had less increase in AQI than Boulder, the model was not statistically significant and did not prove this connection was solely because of policies. Additionally, while the time series does indicate fluctuations in ozone, it does not show a general decreasing trend and also does not take into account other variables such as wildfires and population changes. However, this type of work is still important and relevant as it helps to determine the significance of policy and which ones are most effective, information that helps to inform future plans to combat climate change [@felzer2005future].

*Introduction:*

Air quality is an increasingly urgent public health concern across many urban centers in the United States, and Fort Collins, Colorado, is no exception.
Despite its reputation for clean air and progressive environmental values, Fort Collins has consistently ranked among the worst cities in the nation for ground-level ozone pollution.
According to recent data, the region frequently exceeds the ozone limits set by the Environmental Protection Agency (EPA), placing residents—especially those with respiratory conditions—at heightened risk.
High levels of ozone have been directly linked to increased rates of asthma, chronic bronchitis, and other respiratory diseases, which can result in more frequent emergency room visits and reduced overall quality of life [@chang2006ozone], [@kowalska2009air]. Ground-level ozone, unlike the protective stratospheric ozone layer, is a harmful air pollutant formed through complex photo-chemical reactions [@horn2024air]. Specifically, it is produced when volatile organic compounds (VOCs) and nitrogen oxides (NOx) react in the presence of sunlight and heat. Of these precursors, NOx is primarily emitted from the combustion of fossil fuels, making vehicle emissions a dominant source in urban environments [@lowi1990method].

With the growing population and vehicular traffic in Fort Collins, the city has faced increasing pressure to manage its emissions and reduce ozone pollution levels.
Recognizing the health and environmental risks posed by ozone pollution, Fort Collins has implemented a series of policy initiatives over the past two decades aimed at curbing vehicle emissions and promoting cleaner transportation alternatives.
These include infrastructure investments to support walking and biking, expanded public transportation networks, and efforts to limit single-occupancy vehicle travel.
Additionally, municipal plans have emphasized land use strategies that encourage compact, transit-oriented development and discourage sprawl.
While these policies demonstrate a strong commitment to environmental stewardship, the extent to which they have actually succeeded in reducing ozone pollution remains unclear.
Understanding the effectiveness of these policies is not only critical for guiding future environmental and public health strategies in Fort Collins, but also provides a valuable case study for other cities grappling with similar challenges [@felzer2005future].

As climate change continues to exacerbate heat levels and increase the frequency of high-ozone days, evaluating the success of local mitigation efforts is more important than ever [@chang2006ozone].
This research seeks to address this knowledge gap by systematically examining ozone trends in relation to the implementation of transportation and emissions-related policies in Fort Collins.
The objective of this study is to assess the impact of vehicle emission-reduction policies on ambient ozone levels in Fort Collins.
Specifically, we aim to identify which policy measures have been most successful in lowering ozone concentrations and whether their implementation corresponds to observable improvements in air quality over time.
By evaluating the relative effectiveness of different interventions, we hope to inform both local and regional decision-makers on best practices for managing urban ozone pollution.

To achieve this, we propose a mixed-methods approach that combines quantitative air quality data analysis with a comprehensive policy review.
Ambient ozone concentration data will be obtained from established air quality monitoring stations in and around Fort Collins, covering a multi-year time frame that spans periods before and after the enactment of specific policy initiatives.
Using time-series analysis, we will identify significant changes or trends in ozone levels and assess whether these coincide with policy milestones.
To control for meteorological influences—such as seasonal temperature fluctuations and wind patterns—we will incorporate relevant environmental variables into our models.
In addition to ground-based data and ozone tracking, we will also be relating those findings to policy action and determining if there are connections between the two.
To strengthen the analytical framework, we will conduct a comparative assessment with the city of Boulder, to see if Lairmer's policies has affected the area significantly or not.
This will help contextualize Fort Collins’ progress and provide a more nuanced understanding of policy effectiveness in diverse regulatory environments.
The policy component of the study will involve a detailed review of municipal documents, transportation plans, city council records, and enforcement reports to compile a timeline of enacted policies.
These will be categorized by type—such as infrastructure investments, regulatory changes, or incentive programs—and evaluated for their scope, implementation rigor, and intended outcomes.
Particular attention will be given to measures that directly target vehicle emissions, such as fuel efficiency standards, carpool incentives, or investments in zero-emission public transportation.
To rigorously assess the causal relationship between policy implementation and changes in ozone levels, we will employ statistical techniques such as difference-in-differences (DiD) modeling.
This approach allows us to compare trends in Fort Collins before and after policy enactments against a control group that did not experience similar interventions.
By isolating the effects of policy from broader environmental or economic changes, we aim to draw more robust conclusions about the actual impact of specific measures similar to research that has been conducted before \[@ruggieri2012aggregate\].

The expected outcome of this research is the identification of the most effective policy or set of policies in reducing vehicle emissions and, consequently, ozone pollution in Fort Collins.
We hypothesize that policies which directly address vehicle usage—such as enhanced public transit options or urban design that reduces car dependency—will correlate with the most significant reductions in ozone concentrations.
Ultimately, these findings can guide future environmental policy not only within Fort Collins but also serve as a blueprint for other municipalities seeking to combat ground-level ozone and protect public health in the face of growing environmental challenges.

*Data/Materials:*

Some of the data we will be working with includes pre-generated data files from the Environmental Protection Agency focusing on ozone levels.
We will need to narrow it down substantially in order to only get data for our area of interest, as well as focus on only the parameters which will be relevant to our project, but will most likely not need to clean the data as the majority of it is already well-ordered.
Additionally we will be using legislative documents in order to better understand the policies we will be analyzing and how they did or did not impact ozone levels in Fort Collins.

Data visualization:

```{r}
library(ggplot2)
library(dplyr)

ozone_data <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/daily_44201_2003.csv")
  
filter_data <- ozone_data %>% 
  filter(`County Name` == "Larimer")

ggplot(data = filter_data, aes(x = `Site Num`, y = `AQI`)) +
  geom_point() +
  labs(title = "AQI in Larimer Sites", x = "Site Number", y = "AQI")
```

Figure 1.
Visualization of Larimer's two EPA sites that collect ozone and AQI data.

```{r}
ggplot(data = filter_data, aes(x = `Date Local`, y = AQI)) +
  geom_col() +
  labs(title = "Larimer AQI Over Time", x = "Date", y = "AQI")
```

Figure 2.
Visualization of AQI in Larimer county over time from January 2003 to January 2004.

*Preliminary methods:*

Our methods will be within R, as for analysis we will be focusing on comparing different models that show ozone over time.
At the moment we have everything needed to carry out this analysis, but will need to start narrowing down which policies we will be focusing on and where we will get our data from.
A potential challenge we could face is if the data is if the EPA does not have data that is as focused on our area as we want; if our data is not detailed enough, then we could potentially struggle to get accurate results.
However, if this is the case then we could get supplemental data from the Colorado Department of Public Health and Environment as they have some air quality datasets available.
These datasets have information for specific places which make them useful for our project, but they don't have ozone data over time which makes them harder to work with.

*Methods*

**2.1 Study Design**

This study utilizes a mixed-methods design, incorporating quantitative data to assess the relationship between local policy interventions and ground-level ozone concentrations in Fort Collins, Colorado.
The primary objective is to evaluate whether transportation-related policies aimed at reducing vehicle emissions have led to measurable changes in ambient ozone levels over time.

**2.2 Study Area and Control Selection**

The primary study area is Larimer, Colorado.
As a control, we selected Boulder, Colorado—a nearby city with different policies surrounding ozone in the atmosphere.
The assumption underlying this choice is that differences in ozone trends between the two cities can be attributed, at least in part, to differing policy regimes.

**2.3 Ozone and Meteorological Data Collection**

Ambient ozone concentration data will be collected from the EPA’s Air Quality System (AQS) database and local monitoring stations operated by the Colorado Department of Public Health and Environment (CDPHE).
Data will be collected over a multi-year time frame, spanning at least 10 years to ensure sufficient pre- and post-policy data points.

**2.4 Policy Timeline Development**

A detailed timeline of policy enactments in Fort Collins will be developed through a review of city council records, transportation master plans, and local sustainability reports.
Each policy will be categorized by type (e.g., infrastructure investment, public transit expansion, emission regulation) and date of implementation.
Boulder’s policy timeline will also be reviewed to ensure that it did not implement similar policies during the study period.

**2.5 Data Analysis**

**2.5.1 Time-Series Analysis**

We will conduct a time-series analysis to examine trends in ozone concentrations over time in both Larimer and Boulder county.
Data will be aggregated to monthly averages to smooth out short-term fluctuations while preserving seasonal patterns.

**2.5.2 Difference-in-Differences (DiD) Analysis**

To estimate the causal impact of Larimer’s policies on ozone levels, we will use a difference-in-differences (DiD) statistical model.
This method compares the change in ozone concentrations in Larimer before and after policy implementation with the corresponding change in Boulder over the same period.
The basic DiD model is as follows:

Yit​=α+β1​Postt​+β2​Treatmenti​+β3​(Postt​×Treatmenti​)+γXit​+ϵit​

Where:

Yit​ = ozone level in city iii at time ttt

Postt​ = indicator for post-policy period

Treatmenti​ = indicator for Fort Collins (treatment group)

Xit​ = vector of control variables (e.g., temperature, wind)

β3​ = estimated treatment effect

*Results*

```{r}
#Set up data
#Load data
data_2003 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data.csv")
data_2004 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(1).csv")
data_2005 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(2).csv")
data_2006 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(3).csv")
data_2007 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(4).csv")
data_2008 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(5).csv")
data_2009 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(6).csv")
data_2010 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(7).csv")
data_2011 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(8).csv")
data_2012 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(9).csv")
data_2013 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(10).csv")
data_2014 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(11).csv")
data_2015 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(12).csv")
data_2016 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(13).csv")
data_2017 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(14).csv")
data_2018 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(15).csv")
data_2019 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(16).csv")
data_2020 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(17).csv")
data_2021 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/ozone_fc_03_21/ad_viz_plotval_data(18).csv")

total_data <- rbind(data_2003, data_2004, data_2005, data_2006, data_2007, 
                    data_2008, data_2009, data_2010, data_2011, data_2012, 
                    data_2013, data_2014, data_2015, data_2016, data_2017,
                    data_2018, data_2019, data_2020, data_2021)
```

```{r}
#Time series analysis
library(lubridate)
total_data$Date <- as.Date(total_data$Date, format = "%m/%d/%Y")
total_data$Year <- year(total_data$Date)

total_data %>%
  select(Date) %>%
  distinct() %>%
  slice_head(n = 10)

ozone_fc <- total_data %>%
  mutate(month = floor_date(Date, "month")) %>%
  group_by(month) %>%
  summarize(mean_ozone = mean(`Daily AQI Value`, na.rm = TRUE))
ozone_ts <- ts(ozone_fc$mean_ozone,
               start = c(year(min(ozone_fc$month)), month(min(ozone_fc$month))),
               frequency = 12)
library(forecast)
autoplot(ozone_ts) +
  labs(title = "Monthly Ozone in Fort Collins", y = "Ozone (ppb)", x = "Year")

```

Figure 3.
Time series analysis of Larimer AQI from the years 2003 to 2021.

```{r}
#Difference-in-Differences (DiD) Analysis
full_data_2003 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/daily_44201_2003.csv")
full_data_2021 <- readr::read_csv("/Users/mirandachin/ESS330/lab07/data/daily_44201_2021.csv")

boulder_2003 <- full_data_2003 %>% 
  filter(`County Name` == "Boulder")

boulder_2021 <- full_data_2021 %>% 
  filter(`County Name` == "Boulder")

larimer_2003 <- full_data_2003 %>% 
  filter(`County Name` == "Larimer")

larimer_2021 <- full_data_2021 %>% 
  filter(`County Name` == "Larimer")

did_data <- rbind(boulder_2003, boulder_2021, larimer_2003, larimer_2021) %>% 
  mutate(`Date Local` = as.Date(`Date Local`),
         Year = year(`Date Local`),
         Month = month(`Date Local`))

#Aggregated to year
did_yearly_data <- did_data %>%
  group_by(`County Name`, Year) %>%
  summarise(AQI = mean(AQI, na.rm = TRUE)) %>%
  ungroup()

did_yearly_data <- did_yearly_data %>%
  mutate(
    Treated = ifelse(`County Name` == "Larimer", 1, 0),
    Post = ifelse(Year == 2021, 1, 0),
    Interaction = Treated * Post)

print(did_yearly_data)

did_model <- lm(AQI ~ Treated + Post + Interaction, data = did_yearly_data)
summary(did_model)

#Aggregated to month
did_monthly_data <- did_data %>%
  group_by(`County Name`, Year, Month) %>%
  summarise(AQI = mean(AQI, na.rm = TRUE)) %>%
  ungroup()

did_monthly_data <- did_monthly_data %>%
  mutate(
    Treated = ifelse(`County Name` == "Larimer", 1, 0),
    Post = ifelse(Year == 2021, 1, 0),
    Interaction = Treated * Post)

did_monthly_model <- lm(AQI ~ Treated + Post + Interaction, data = did_monthly_data)
summary(did_monthly_model)

#Data visualization
avg_trend <- did_monthly_data %>%
  group_by(Month, Year, Treated) %>%
  summarise(mean_AQI = mean(AQI, na.rm = TRUE)) %>%
  mutate(
    Time = as.Date(paste(Year, Month, "01", sep = "-")))


ggplot(avg_trend, aes(x = Time, y = mean_AQI, color = factor(Treated))) +
  geom_line(size = 1.2) +
  geom_vline(xintercept = as.Date("2021-01-01"), linetype = "dashed", color = "black") +
  labs(
    title = "DID Model with AQI Over Time",
    x = "Time",
    y = "Monthly Average AQI",
    color = "Treated") +
  theme_minimal()

```

Figure 4.
Difference-in-Differences model visualization comparing only the AQI of years 2003 and 2021 of Larimer County and Boulder county, labeled 0 and 1 respectively.

The time-series analysis of monthly ozone concentrations in Fort Collins shows strong seasonal patterns, with recurring peaks in the summer months (Figure 3).
Although some fluctuations occur year to year, there is no clear or sustained downward trend following the 2015 implementation of local transportation policy interventions.

In the DiD monthly model in Figure 4, Boulder's ozone AQI went up by 8.4 units while Larimer's ozone was smaller by 6.3 units, indicating that Larimer had a smaller increase in ozone.
However, the p-value is 0.741, which is not scientifically significant; this means that the model cannot explain variations in the data and these differences could be from random error.

*Discussion*

Looking at the Difference-in-Differences (DiD) model and Figure 4, the trends between Larimer and Boulder are very similar.
However, the actual model itself resulted in a p-value above 0.05, indicating that the model did not yield statistically significant results and cannot explain the data accurately.
This implicates that although these policies have had a positive impact on air quality in Larimer County, we still cannot say for sure if policy is the sole reason for improvement.
Looking at the monthly AQI of Larimer in the time series visualization of Figure 3, it implies that ozone AQI has continued to experience fluctuations both with and without new policy.

Several factors may explain this inconclusive result.
First, the policies implemented in Fort Collins may not have been strong or sustained enough to produce a detectable change in ozone levels within the timeframe studied.
Ozone formation is influenced by a complex set of variables, including temperature, regional air currents, and emissions from neighboring areas, all of which may dilute the local impact of policy efforts [@kumari2018critical].
Additionally, ozone pollution is a secondary pollutant, meaning it forms from precursor chemicals rather than being directly emitted.
This makes reductions in ozone slower and more indirect than reductions in primary pollutants like NO₂ or PM₂.₅ [@zhan2018driving].

Another possibility is that the policy interventions overlapped with other regional or statewide efforts, making it difficult to isolate the effect of Fort Collins' actions alone.
The DiD approach assumes no other major interventions or shocks affected one city and not the other during the study period, but this assumption may not hold perfectly given Colorado’s broader air quality regulations and environmental shifts.

These results relate to the broader field of science by displaying how models can be used to find patterns and trends in data over decades of time.
It also connects to the need for accurate data and models in order to make decisions about environmental policies and changes on a local scale.
For the future, similar projects can work to include other variables beyond policy changes, as well as dive deeper into differences depending on if data is collected on a different scale than county, such as by local communities, state, or other.

*Conclusion*

This study aimed to evaluate whether transportation-related policy interventions in Fort Collins led to measurable improvements in ozone-related air quality.
While time-series trends suggest minor changes and the DiD visualization shows some divergence between Fort Collins and Boulder, the lack of statistically significant results means we cannot confidently attribute these differences to the policies in question.

Although the findings are inconclusive, they highlight the complexities of air quality governance and the challenges in isolating the effects of local interventions.
Future work should incorporate longer time horizons, additional control cities, and more detailed meteorological and emissions data.
A broader range of methods—such as synthetic controls or spatial regression—could also strengthen causal inference.
Importantly, the lack of significant results does not mean the policies failed, but rather that their effects are difficult to detect using the methods and data currently available.
